### 2017-01-01 paper/paris17/paper.tex
personal review

Propagating wavefields using explicit finite difference method is the kernel of reverse time migration (RTM) and high-end velocity algorithms in seismic applications. In recent decades there has been a significant increase of interest in the seismic exploration community to reconstruct the image of the underground in larger regions and higher resolutions, especially in the elastic media. As a result the optimizing methods to improving the performance of the wavefield propagation are in great demands.

@@

One approach to improve the performance of modeling is to recognize that the earth attenuates seismic energies, in which case, we can ignore higher and higher frequencies and still obtain an accurate image\cite[]{Clapp.sep.111.bob3} when we push the wavefield down in depth. By using coarser sampling at large time steps,  the performance for downward continuation based approaches which are done frequency by frequency can be significantly improved.

The concept of the constant Q model introduced by \cite{Kjartansson.sep.23} are used to approximate the attenuation. However, the formulation for computing the time derivative for the constant Q model proposed by Podlubny requires to store wavefields from previous values to present time \cite[]{podlubny1998fractional}, which is impractical for real seismic modeling in terms on memory capacities on modern computers. Zhu developed a constant-Q wave equation using fractional Laplacian operators to avoid the need of storing the past values of the wavefields \cite[]{zhu2014theory}. However, the proposed method is for 2D viscoelastic case and the computation of fractional Laplacian is expensive compared with the conventional FD methods.

Meanwhile, the Nvidia Graphic Processing Units (GPUS)  have achieved great successes in many domains. In seismic application, GPUs are employed as high-performance computing accelerators for key seismic modeling, imaging and inversion kernels due to their high memory bandwidth and a large number of processing units \cite[]{ohmer2005implementation,he2015gpu}.

This work proposes a high performance design of accelerating the wavefield propagation in 3D elastic media by approximating the constant Q propagation on multiple Nvidia GPUs. We first propose the Q approximation formulation by extending the formulation from the 2D viscoelastic to the 3D viscoelastic case where the fractional Laplacian is approximated to the conventional Laplacian and the dispersion term is ignored.  Optimization strategies from different aspects (memory, occupancy, and overlapping) are performed to form a high performance modeling kernel on multiple GPUs. To further improve the performance, a couple of tricks are also applied to reduce the total computation. Combining all the optimization schemes, we can achieve a significant speedup of the 3D elastic wave propagation by 60 to 200 times over a highly optimized 24-core CPU solution.

@@

Our theory is built on the foundation of \cite[]{zhu2014theory}. It is a time-domain differential equation for modeling seismic wave propagation in constant-Q viscoelastic media based on fractional spatial derivatives, specifically Laplacian differential operators of fractional order.  The stress–strain relation is derived from the classical equation expressed in terms of fractional time derivatives. The formulation has the advantage of not requiring additional field variables that increase the computer time and storage significantly \cite[]{zhu2014theory}.

We first extend the formulation of \cite[]{zhu2014theory} from the 2D viscoelastic case to the 3D viscoelastic case. Note that the Laplacian differential operators of fractional order bring extra computational costs to the elastic modeling, we approximate it with the conventional Laplacian operators, which significantly reduces the computing complexities. Also, we further reduce the computational costs by ignoring the dispersion terms because it make little effects compared to the attenuation of the signals. As a result, our equations are then

@@

The reason why our constant Q formulation is relatively computationally economic are: (1) The fractional Laplacian is approximated to the conventional Laplacian. (2) The dispersion term is ignored. (3) We preserve the original workflow of elastic modeling while only adding a few terms in the calculation of stress tensors from train tensors.

@@

The reason why GPUs outperforms CPUs significantly on many scenarios is that the architecture of GPUs favors throughput of many data-parallel tasks over the latency of a single thread. We have an efficient parallel design on the GPU platform with three main strategies.

Firstly, we fully utilize the different hierarchical memory of the GPU. We keep a set of constant variables such as the derivative coefficients in the constant memory and the read-only cache. As the 64KB first-level cache can be configured as different combinations between L1 cache and shared memory, we configure it dynamically based on different computing scenarios. For the stencil computation, we configure the first-level cache by setting the sizes of shared memory and L1 cache to be 48 KB and 16 KB, respectively, which achieves a high computation performance based on P. Micikevicius’ 3D stencil design. For others, we prefer to configure the first-level cache as L1 cache if we don't use shared memory explicitly.

Secondly, we carefully tune the number of registers per thread to maximize the occupancy of GPUs. Since the access pattern of GPU global memory in the stencil kernel is irregular, the more threads are busy, the less latencies it is for memory access. However, as the number of registers is limited in a streaming multiprocessor (SM), increasing the number of registers per threads will decrease the number of active warps. So it is important to have a strategy to balance them. We first remove the intermediate variables to reduce the number of registers per threads. Then we dynamically tracking the performance by adjusting the block size and the \emph{maxrregcount} compiler option.

Thirdly, we design a computation and communication overlapping scheme to maximize the concurrency of different resources. The entire wavefield domain is divided into 4 subdomain, each assigning to a GPU shown in Figure \ref{fig:domain-decomposition}. Note that we need ghost cells to store and exchanges the halos, denoted by the arrows. In each subdomain, we further divide it into three parts, the halo part, the outer part and the inner part shown in Figure \ref{fig:task-decomposition}. The computation of inner and outer parts are assigned to different CUDA streams so that they are executed concurrently. When the computation of the outer part is finished, the GPU starts exchanging halos with its neighbors. In this way, the computation of the inner part can hide the communication shown in Figure \ref{fig:overlap}.

@@

Apart from the GPU-based optimization, we also apply another two approaches to further improve the performance.  After we taking the energy attenuation into consideration, we can ignore the frequencies that have attenuated. As a result, the grid cells and the time steps get larger as the wavefield propagate in time. We can achieve significant speedup when we resample the velocities and wavefields to coarser grids, especially recording the longer time step. Using the above approach, the early time steps dominate the computation. We eliminate the bottleneck by applying another trick that we recognize that it is not necessary to propagate the wavefields significantly away from the source at early times. We eliminate the update of the wavefields where the waves have not reached so far.

@@

\subsection{Q Effects}
Experiments for demonstrating the functionality of applying the Q effects as well as  the performance with different optimization tricks are carried out in this paper. Figure \ref{fig:with-attenuation} shows a slice of the z-component of the 3D wavefield using standard elastic wave equation at time 0.9 seconds while the Figure \ref{fig:without-attenuation} shows the wavefields with the constant Q effect ($Q_s=Q_p=20$) added. The wave front in Figure \ref{fig:with-attenuation} is weaker. The spectrum of the wavefields gives a better illustration, which is shown in Figure \ref{fig:spectrum}.

@@

We propagate the wavefield in the 3D domain of size $700\times700\times700$ for 6000 time steps to measure the performance. Figure \ref{fig:gpu-speedup} shows the performance speedup of 1-4 K40 GPUs over one 24-core CPU node. The GPU design among 4 K40 accelerator cards can improve the performance of the computational part by up to 96 times, and improve the overall performance by up to 63 times. Note that the CPU design is also tuned by using multi-threading and vectorization. Figure \ref{fig:q-speedup} shows the speedup factor, which is  defined as the number of grid cells times time steps, as a function of propagation time for different Q values. We can see that the longer the time records, we more significant performance speedup we achieve.  Combing the two tricks as well as the GPU-based optimizations, we can achieve a performance speedup over 60 to 200 times as a function of Q.

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \centering
        \includegraphics[height=1.5in]{./fig/speedup.eps}
        \caption{Speedup with GPU optimizations}
        \label{fig:gpu-speedup}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \centering
        \includegraphics[height=1.5in]{./fig/speedup_q.eps}
        \caption{Speedup of different Q value}
        \label{fig:q-speedup}
    \end{subfigure}
    \caption{Performance speedup}
\end{figure}

@@

This work manages to boost the performance the wave propagation in 3D elastic scenarios by approximating the Q propagation and using GPU platforms. We first extends the constant-Q formulation from the 2D viscoelastic to 3D viscoelastic case. Different optimization techniques on GPUs are then described for a high performance modeling kernel. We also propose a set of tricks to reduce the computation to further reduce computations. The experimental results show that we can achieve significant performance speedup by 60 to 200 times over the CPU-based solution.

>>>>

Propagating wavefields using the explicit finite difference method is the kernel of reverse time migration (RTM) and high-end velocity algorithms in seismic applications. In recent decades there has been a significant increase of interest in the seismic exploration community to invert the image of the subsurface in larger regions and higher resolutions in the elastic media, which brings tremendous computing challenges. As a result the optimizing methods for improving the performance of the wavefield propagation are in great demands.

@@

One approach to improve the performance of modeling is to recognize that the Earth attenuates seismic energy. We can ignore higher and higher frequencies and still obtain an accurate image\cite[]{Clapp.sep.111.bob3} when we push the wavefield down in depth. By using coarser sampling at large time steps based on the stability and dispersion conditions, we can reduce the computational cost for wavefield propagation.

The concept of the constant Q model introduced by \cite{Kjartansson.sep.23} are used to approximate the attenuation. However, the formulation for computing the time derivative for the constant Q model requires to store wavefields from previous values to present time \cite[]{podlubny1998fractional}, which is impractical for real seismic modeling in terms of memory capacities on modern computers. Zhu developed a constant-Q wave equation using fractional Laplacian operators which avoids the need of storing the past values of the wavefields \cite[]{zhu2014theory}. However, the proposed method is for 2D viscoelastic case and the computational cost of fractional Laplacian is still expensive compared with the conventional Laplacian.

Meanwhile, the Nvidia Graphic Processing Units (GPUS)  have achieved a great success in many domains. In seismic applications, GPUs are employed as high-performance computing accelerators for key seismic modeling, imaging and inversion kernels due to their high memory bandwidth and a large number of processing units \cite[]{he2015gpu}.

This work proposes a high performance design of accelerating the wavefield propagation in 3D elastic media by approximating the constant Q propagation on multiple Nvidia GPUs. We first propose the Q approximation formulation by extending the formulation from the 2D viscoelastic to the 3D viscoelastic case where the fractional Laplacian is approximated to the conventional Laplacian and the dispersion term is ignored.  Optimization strategies from different aspects (memory, occupancy, and overlapping) are performed to form an efficient kernel on multiple GPUs. To further improve the performance, a set of tricks are also applied to reduce the computation. Combining all the optimization schemes, we can achieve a significant speedup of 60 to 200 times over a highly optimized 24-core CPU solution for the 3D elastic wavefield propagation.

@@

Our theory is built on the foundation of \cite{zhu2014theory}. The time-domain differential equation is based on the Laplacian differential operators of fractional order, which is able to model seismic wave propagation in constant-Q viscoelastic media. The stress-strain relation is derived from the classical equation expressed in terms of fractional time derivatives. The formulation has the advantage of not requiring additional field variables so that it saves the computing time and storage significantly \cite[]{zhu2014theory}.

We extend the formulation of \cite{zhu2014theory} from the 2D viscoelastic case to the 3D viscoelastic case. Note that the Laplacian differential operators of fractional order bring extra computational costs to the elastic modeling, we approximate it with the conventional Laplacian operators, which significantly reduces the computing complexities. Also, we further reduce the computational costs by ignoring the dispersion terms because it make little effects compared to the attenuation of the signals. As a result, our equations are as follows.

@@

The reasons why our constant Q formulation is relatively computationally economic are as follows. (1) The fractional Laplacian is approximated to the conventional Laplacian. (2) The dispersion term is ignored. (3) We preserve the original workflow of elastic modeling while only adding a few terms when computing stress tensors from train tensors.

@@

The reason why GPUs outperforms CPUs significantly on many scenarios is that the architecture of GPUs favors the throughput of many data-parallel tasks over the latency of a single thread. We have an efficient parallel design on the multiple-GPU platform with three main strategies.

Firstly, we fully utilize the hierarchical memory of the GPU. We keep a set of constant variables such as the derivative coefficients in the constant memory and the read-only cache. As the 64KB first-level cache can be configured as different combinations between L1 cache and shared memory, we configure it dynamically in different scenarios. For the 8th-order stencil computation, we configure the first-level cache by setting the sizes of shared memory and L1 cache to be 48 KB and 16 KB respectively, which achieves great performance based on the 3D stencil design proposed by \cite{micikevicius20093d}. For others, we prefer to configure the first-level cache as L1 cache where we don't use shared memory explicitly.

Secondly, we carefully tune the number of registers per thread to maximize the occupancy of GPUs. Since the access pattern of GPU global memory in the stencil kernel is irregular, the more threads there are, the less latency it gains for memory access. However, as the number of registers is limited in a streaming multiprocessor (SM), increasing the number of registers per thread will decrease the number of active warps. So it is important to have a strategy to balance them. We first remove the intermediate variables to reduce the number of registers per thread. Then we dynamically track the performance by adjusting the block size and the \emph{maxrregcount} compiler option.

Thirdly, we design a computation and communication overlapping scheme to maximize the utilization of different kinds of resources. The entire wavefield domain is divided into 4 subdomains, each assigning to a GPU shown in Figure \ref{fig:domain-decomposition}. Note that we need ghost cells to store and exchange the halos, denoted by the arrows. In each subdomain, we further divide it into three parts, the halo part, the outer part and the inner part as shown in Figure \ref{fig:task-decomposition}. The computation of inner and outer parts are assigned to different CUDA streams so that they are executed concurrently. When the computation of the outer part is finished, the GPU starts exchanging halos with its neighbors. In this way, the computation of the inner part can hide the communication time perfectly, shown in Figure \ref{fig:overlap}.

@@

Apart from the GPU-based optimization, we also apply another two approaches to further improve the performance.  (1) After taking the energy attenuation into consideration, we can ignore the frequencies that have attenuated. Therefore, the grid cells and the time steps can get larger as the wavefields propagate in time. We can achieve significant speedup when we resample the velocities and wavefields to coarser grids, especially in longer time steps. Using the above approach, the early time steps dominate the computation. (2) We eliminate the bottleneck by applying another trick that it is not necessary to propagate the wavefields significantly away from the source at early times. So the update of the wavefields where the waves have not reached so far is eliminated.

@@

\subsection{Q Effects}
Experiments for demonstrating the functionality of applying the Q effect as well as  the performance with different optimization strategies are presented. Figure \ref{fig:with-attenuation} shows the z-x pane of the z-component of the 3D wavefield using standard elastic wave equation at time 0.9 second while the Figure \ref{fig:without-attenuation} shows the wavefields with the constant Q ($Q_s=Q_p=20$) added at the same time step. We can see that the wavefields are nearly identical except that the wavefront in Figure \ref{fig:with-attenuation} is weaker, which is reasonable owning to the energy attenuation. The spectrums of the wavefields give a better illustration for the attenuation, which is shown in Figure \ref{fig:spectrum}.

@@

We propagate the wavefield in the 3D domain of size $700\times700\times700$ for 6000 time steps to measure the performance. Figure \ref{fig:gpu-speedup} shows the performance speedup of 1-4 K40 GPUs over one 24-core CPU node. The GPU design among 4 K40 GPUs can significantly improve the performance of the computational part by 96 times, and improve the overall performance by 63 times. Note that the CPU design is also well tuned by using multi-threading and vectorization. Figure \ref{fig:q-speedup} shows the speedup as a function of propagation time for different Q values. We can see that we obtain a great performance gain at early and late time because we restrict the computing region at early time and ignore the high frequencies at late time. The longer the time records, the more significant speedup we achieve.  Combing the all optimization schemes as well as the GPU-based optimizations, we can achieve a performance speedup of 60x to 200x as a function of Q.

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \centering
        \includegraphics[height=1.5in]{./fig/speedup.eps}
        \caption{Speedup with GPU optimizations}
        \label{fig:gpu-speedup}
    \end{subfigure}%
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \centering
        \includegraphics[height=1.5in]{./fig/speedup_q.eps}
        \caption{Speedup as different Q value}
        \label{fig:q-speedup}
    \end{subfigure}
    \caption{Performance result of GPU optimizations and energy attenuation}
\end{figure}

@@

This work manages to boost the performance the wavefield propagation in 3D elastic scenarios by approximating the Q propagation and the using multi-GPU platform. We first extend the constant-Q formulation from the 2D viscoelastic to the 3D viscoelastic case. Different optimization techniques on GPUs are then described for an efficient modeling kernel. We also propose a set of schemes to reduce the computation to further improve performance. The experimental results show that we can achieve significant performance speedup of 60 to 200 times with 4 GPUs over the CPU-based solution as a function of Q.
